---
title: "The Generative Texture Model"
author: "Sean T. Bresnahan and Charis Xiong"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The Generative Texture Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, comment = "#>"
)
set.seed(1)
library(iconic)
```

# Two simulation modes

ICONIC uses two distinct simulation modes that serve different purposes:

1. **Benchmark mode** (`generate_toy_data`): a structural data-generating
   process with parametric Gaussian noise and no learned texture. The
   ground truth (NDE, NIE, confounding strength, instrument strength) is a
   closed-form function of the parameters. This mode is used to validate
   the estimators under controlled confounding scenarios with exactly known
   ground truth.

2. **Data-calibrated mode** (the generative texture pipeline): trains a
   texture model on the user's own data so that the sensitivity and
   prospective analyses are calibrated to realistic covariate, outcome,
   and mediator distributions. The ground truth is still imposed by the
   structural skeleton, but the nuisance texture (marginals, correlations,
   covariate structure) is learned from the data.

This separation is deliberate: estimator validation requires a transparent,
reproducible data-generating process whose parameters are known exactly,
while the user-facing workflow benefits from realistic data texture even
at the cost of a learned noise model.

# The hybrid GAN + copula architecture

The texture model is a hybrid of two independently trained components,
each chosen to match the dimensionality and structure of its target block:

- **Sample-level GAN**: an MLP generator and discriminator trained with a
  non-saturating BCE-with-logits objective, Adam optimization
  ($\beta_1 = 0.5$), and one-sided label smoothing. This GAN learns the
  joint distribution of the per-sample exposure level, outcome level,
  mediator level, and encoded covariates (one row per sample). A GAN is
  used here because the block contains mixed continuous and categorical
  variables whose nonlinear interactions and mutual-exclusivity constraints
  are more naturally preserved by a flexible generator than by a Gaussian
  copula.

- **Feature-level Gaussian copula**: learns the full joint distribution of
  the mediator ($M$) panel. Each feature's marginal is fit by empirical
  CDF, or by a parametric family (normal, log-normal, gamma, or beta,
  selected by AIC) when it passes a Kolmogorov-Smirnov test at $p > 0.05$.
  Cross-feature dependence is captured by the correlation matrix of the
  normal-score-transformed marginals.

Neither component learns the causal effect. The learned texture is injected
into the same structural causal model as the benchmark mode, but drawn
independently of the latent confounders $U$ so it cannot open a hidden
backdoor. Because the generative model never models the joint of
$(Z, M, Y)$ and only learns the nuisance texture, the ground-truth effect
is guaranteed by construction.

# Training the texture model

The texture model requires `torch`. Verify your installation first:

```{r texture-check}
check_torch_setup()
```

If this returns `FALSE`, install torch with
`install.packages("torch")` and run `torch::install_torch()`.

We simulate a small panel and train the texture model briefly:

```{r texture-data}
dat <- generate_toy_data(
  n = 100, n_features = 5, seed = 1,
  phi = 0.8, separate_U = TRUE, omega_1 = 0.7, omega_2 = 0.7
)

idat <- iconic_data(
  Z = dat$Z, Y = dat$Y, M = dat$M,
  G = dat$G, Gm = dat$Gm,
  W1 = dat$W1, W2 = dat$W2,
  covariates = dat$synthetic_data
)
```

```{r texture-train}
input <- load_real_input_data(
  Z_matrix = matrix(dat$Z, nrow = 1),
  Y_matrix = t(dat$Y), M_matrix = t(dat$M),
  W_matrix = t(dat$W), covariates_df = dat$synthetic_data
)

gan <- train_gan_on_real_data(
  input$gan_training_data,
  feature_correlations = input$feature_correlations,
  feature_texture = input$feature_texture,
  epochs = 5, seed = 1, verbose = FALSE
)

gan
```

In practice, use 100–300 epochs for a production-quality texture model.
The `epochs = 5` setting here is for speed only.

# Attaching the texture model

Once trained, attach the texture model to the `iconic_data` object so that
every downstream function (`iconic_sensitivity()`, `iconic_prospect()`)
reuses it without retraining:

```{r texture-attach}
idat$trained_gan <- gan
```

You can also pass a pre-trained model via the `trained_gan` argument to
`iconic_data()` at construction time.

# When to use each mode

| Mode | When to use | Ground truth | Texture |
|---|---|---|---|
| Benchmark (`generate_toy_data`) | Estimator validation, benchmarks | Known exactly (closed-form) | Parametric Gaussian |
| Data-calibrated (GAN + copula) | Sensitivity, prospective analysis | Imposed by structural skeleton | Learned from your data |

The sensitivity and prospective analyses benefit from realistic texture
because estimator performance depends on the empirical distribution of the
analysis data. The benchmark mode is for transparent, reproducible
validation where the ground truth must be known exactly.

# Further reading

- **Sensitivity vignette**: `vignette("iconic-sensitivity")` — how the
  trained texture model is used in the sensitivity sweep.
- **Function reference**: `?train_gan_on_real_data` for training arguments,
  `?check_torch_setup` for the torch verification helper.
