---
title: "ICONIC: Causal Model Selection with Genetic Instruments and Negative Controls"
author: "Sean T. Bresnahan and Charis Xiong"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ICONIC Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
subtitle: Model selection, simulation, and sensitivity analysis for observational omics
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, comment = "#>"
)

set.seed(1)
library(iconic)
```

# What this package is for

`iconic` is a workbench for causal inference in observational omics studies
where unmeasured confounding biases ordinary regression. The motivating
application is estimating the causal effect of gestational diabetes mellitus
(GDM) on placental transcription, but the framework generalizes to any study
with genetic instruments and/or negative controls.

The package provides:

1. **Eight causal estimators** (including an unadjusted reference), each
   leaning on a different identifying assumption — a genetic instrument, a panel
   of negative-control outcomes, or both.
2. A **generative pipeline** that trains a hybrid texture model — a
   sample-level GAN for covariates/exposure/outcome plus a feature-level
   Gaussian copula for the mediator panel — on *your real dataset* and
   produces realistic in-silico data with a known ground-truth effect, so you
   can benchmark estimators on data that looks like yours.
3. A **sensitivity analysis** that sweeps confounding scenarios and reports
   which estimator to trust and whether your negative controls remain valid.
4. **Mediation estimators** for natural direct (NDE) and natural indirect
   (NIE) effects, with benchmarking under mediator-outcome confounding.
5. **Pleiotropy and instrument-strength diagnostics** for benchmarking under
   exclusion-restriction violations and weak instruments, plus empirical
   negative-control validity screens with FDR control.

The statistical theory — structural causal model, identification conditions,
estimator derivations — is documented in the accompanying manuscript. This
vignette focuses on how to use the package.

---

# Model selection on your own data

ICONIC provides a four-step workflow for choosing the right causal estimator
for your data: diagnose, estimate, stress-test, and recommend.

1. **Diagnose** (`iconic_diagnose()`) — check instrument strength (partial
   F-statistics), negative-control validity (A1/A2/A2' screens with BH-FDR
   control), and path completeness. Returns an eligibility table for all
   eight estimators.
2. **Estimate** (`iconic_estimate()`) — fit all eligible estimators on your
   real data. Returns per-feature (and per-mediator) NDE, NIE, SE, and
   p-values.
3. **Stress-test** (`iconic_sensitivity()`) — map the degradation surface:
   how each estimator's bias changes as instrument-independence assumptions
   are violated. The surface is calibrated to your sample size, covariate
   texture (via an auto-trained hybrid GAN + copula model), and confounding
   parameters (which can be inferred from the data via `infer_confounding()`).
4. **Recommend** (`iconic_recommend()`) — rank eligible estimators by
   identification strength (tier A: identified, B: negative-control, C:
   instrument-based, D: bias reference) and robustness. Returns the
   top-ranked estimator with justification.

## Step 1: Standardize your data

All matrices are stored as features x samples (rows = features, cols =
samples). If you pass samples x features, ICONIC auto-transposes.

```r
library(iconic)

# Example with all inputs (exposure, outcome, mediator, instruments, NCs)
data <- iconic_data(
  Z = my_exposure,           # numeric vector, one per sample
  Y = my_outcomes,           # features x samples matrix (or vector)
  M = my_mediator,           # features x samples matrix (or vector)
  G = my_prs,      # genetic instrument (e.g. PRS)
  Gm = my_eqtls,             # mediator instrument (e.g. cis-eQTLs)
  W = my_negative_controls,  # features x samples matrix
  covariates = my_covariates # data frame (n rows)
)
print(data)
```

When only `W` is supplied (single-panel NCs), ICONIC sets `W1 = W2 = W` for
backward compatibility with the single-panel estimators (COCA, PGC). To use
the path-specific estimators (PGC2, PGC2Gm), supply `W1` and `W2` separately.

You can optionally attach a pre-trained texture model via
`trained_gan = my_gan`. This avoids retraining when the same data is used
across multiple workflow steps (e.g. sensitivity + prospective analysis). If
omitted, `iconic_sensitivity()` and `iconic_prospect()` will auto-train a
generative model from your data on first use.

## Step 2: Diagnose — which estimators are valid?

```r
diag <- iconic_diagnose(data)
print(diag)
# Reports instrument strength (partial F for G and Gm),
# NC validity screens (A1, A2, A2'), completeness, and
# an eligibility table for all 8 estimators.
```

`iconic_diagnose()` computes the partial F-statistic for every instrument
and every mediator in the panel (not just the first). The full distributions
are stored at `diag$instrument_strength$F_G` and
`diag$instrument_strength$F_Gm` (numeric vectors, one entry per
instrument/mediator), so you can inspect the spread of instrument quality
across the panel — e.g. a histogram of eQTL strengths, or the fraction
clearing a given threshold.

### Panel-level instrument gating

By default, eligibility uses the legacy scalar rule: an instrument-dependent
method is eligible if the **median** partial F across the panel meets
`min_f` (Stock-Yogo threshold, default 10). This preserves backward
compatibility for single-instrument analyses.

For mediator panels with heterogeneous instrument quality (e.g. cis-eQTLs
whose strength varies across transcripts), the scalar median can be
misleading — a strong median may hide a large fraction of weak
instruments. The `g_threshold` and `gm_threshold` arguments let you gate
eligibility on the **distribution** instead:

```r
# Eligible if >= 50% of transcripts have F_Gm >= 10
diag <- iconic_diagnose(data,
  gm_threshold = list(E = 0.5, R = 10))
print(diag)
# Gm (mediator instrument): median F = 4.5, 16% of 9300 transcripts >= 10
#   (BELOW E) (E threshold: 50%)
# Eligible estimators: 3/8  (UNADJ, DIRECT, COCA)
```

- **R** = the per-transcript F cutoff (how strong is "strong enough" for one
  instrument?).
- **E** = the fraction of transcripts that must clear R for the method to be
  eligible at all (how much of the panel needs to be well-instrumented?).

The decision is `weak = (mean(F >= R) < E)`. When the threshold is not met,
Gm-dependent methods (IV2SLS2, PGC2Gm) are marked ineligible. The same
logic applies symmetrically to G via `g_threshold` (useful when G is a
matrix of multiple instruments). When a threshold is `NULL` (the default),
the legacy scalar rule (median F vs `min_f`) is used.

### Inspecting the distribution

```r
fgm <- diag$instrument_strength$F_Gm   # length-9300 numeric vector
summary(fgm)
#    Min.  1st Qu.  Median    Mean  3rd Qu.    Max.
#  0.000   0.800    4.500    9.200   12.400  1850.0

# Pass fractions at common thresholds
for (R in c(3, 5, 10, 20))
  cat(sprintf("  F_Gm >= %2d: %5.1f%%\n", R,
              100 * mean(fgm >= R, na.rm = TRUE)))
```

This lets you choose E and R informed by the actual distribution rather than
guessing.

The eligibility rules are:

| Estimator | Requires |
|-----------|----------|
| UNADJ | always eligible |
| DIRECT | G + W |
| COCA | W, valid NCs (A2 screen) |
| IV2SLS | G, $F_G$ meets threshold, W |
| PGC | G, $F_G$ meets threshold, completeness satisfied |
| IV2SLS2 | G + Gm + W, $F_G$ and $F_{Gm}$ meet thresholds |
| PGC2 | G + W1/W2, $F_G$ meets threshold, path completeness |
| PGC2Gm | G + Gm + W1/W2, $F_G$ meets threshold, path completeness |

"Meets threshold" means either median F $\geq$ `min_f` (legacy scalar, default)
or the E/R panel rule (when `g_threshold`/`gm_threshold` is supplied).

## Step 3: Estimate — fit all eligible estimators

```r
est <- iconic_estimate(data, diagnosis = diag)
head(est)
# Per-feature NDE, NIE, SE, p-value
# for each eligible method.
```

If `diagnosis` is supplied, only eligible estimators are run. You can
override with `methods = c("IV2SLS", "PGC")` to run specific estimators
regardless of eligibility.

### Per-transcript weak-instrument gate (`min_f`)

Even when a method is eligible, individual transcripts whose instrument is
too weak receive `NA` rather than a biased estimate. The per-transcript
cutoff is controlled by `min_f`, which is **inherited from the diagnosis**
when not explicitly set:

```r
# Diagnose with min_f = 3; estimate inherits it automatically
diag <- iconic_diagnose(data, min_f = 3)
est  <- iconic_estimate(data, diagnosis = diag)   # uses min_f = 3

# Override with an explicit value
est  <- iconic_estimate(data, diagnosis = diag, min_f = 10)
```

When no diagnosis is supplied and `min_f` is not set, it defaults to 10
(Stock-Yogo). Transcripts with F < `min_f` return `NA` for NDE/NIE; those
that pass get an estimate. This is separate from the eligibility decision:
eligibility gates whether the method runs at all; `min_f` gates which
transcripts get non-NA results.

### Forcing all methods (`run_all`)

```r
# Run every method whose required data exists, ignoring F thresholds
est <- iconic_estimate(data, run_all = TRUE, min_f = 10)
```

`run_all = TRUE` overrides diagnosis-based eligibility, running every method
for which the required inputs (G, Gm, W, etc.) are present. The per-transcript
`min_f` gate still applies. This is useful for exploratory analysis where you
want to see all estimates and judge instrument quality yourself.

### Standard errors and p-values (`se_method`)

ICONIC offers three methods for computing the standard error and p-value of
the natural indirect effect (NIE = $\alpha_M \times \beta_M$):

| `se_method` | NIE SE | NIE p-value | Speed | When to use |
|---|---|---|---|---|
| `"delta"` (default) | Delta-method (product formula) | Sobel / Wald z-test | Fast | General-purpose; conservative under sparse signals |
| `"bootstrap"` | Nonparametric bootstrap SD (`n_boot` resamples) | Wald z-test with bootstrap SE | Slow | When delta-method normality is in doubt; CI construction |
| `"composite"` | Delta-method (unchanged) | Huang (2019) JT-comp composite null test | Fast | When signals are sparse and Sobel is too conservative |

```r
# Default: delta-method SE + Sobel p-value
est <- iconic_estimate(data, diagnosis = diag, se_method = "delta")

# Bootstrap SE (slower; use for CI construction)
est <- iconic_estimate(data, diagnosis = diag, se_method = "bootstrap",
                       n_boot = 500)

# Composite null p-value (JT-comp; higher power under sparse signals)
est <- iconic_estimate(data, diagnosis = diag, se_method = "composite")
```

**Why three options?** The NIE is a *product* of two coefficients
($\hat\alpha_M$ and $\hat\beta_M$). The null $H_0: \alpha_M \beta_M = 0$ is
*composite* — it holds if either coefficient is zero. The Sobel (delta) test
approximates the product as normal, which is only valid when one coefficient
is nonzero and the other is zero (the "partial null" cases). When *both* are
zero (the "point null" or sparse-signal case), the product follows a normal
*product* distribution, not a normal, making the Sobel test conservative.

The **JT-comp** test (Huang 2019) accounts for all three null cases via a
closed-form p-value based on the normal product distribution CDF:

$$\hat p_{\text{comp}} = F\!\left(\frac{ab}{\sqrt{\widehat{\text{Var}}(a)}}\right)
+ F\!\left(\frac{ab}{\sqrt{\widehat{\text{Var}}(b)}}\right) - F(ab)$$

where $a = \hat\alpha_M / \text{SE}(\hat\alpha_M)$ and
$b = \hat\beta_M / \text{SE}(\hat\beta_M)$ are the standardized
z-statistics, $\widehat{\text{Var}}(a)$ and $\widehat{\text{Var}}(b)$ are
their variances across the collection of tests, and $F(z)$ is the CDF of the
standard normal product distribution. The variances are estimated per
estimator across all tests sharing that estimator — i.e., across mediators
(and outcome features, when present) within each method. When fewer than 5
tests are available, the point-null value $\text{Var} = 1$ is used.

**When `se_method = "composite"`:**

- `NIE_se` is unchanged (delta-method SE, used for CI construction).
- `NDE_p` is unchanged (the NDE is a single coefficient, not a product).
- Only `NIE_p` is replaced by the JT-comp composite p-value.
- The output includes `var_a` and `var_b` columns for transparency.

**Limitations of the composite test:**

- The JT-comp test cannot distinguish the partial-null case
  $H_0^{(2)}: \alpha \neq 0, \beta = 0$ from the alternative when
  $\alpha_M$ is large, because the product $ab$ can be large even when
  $b \sim N(0, 1)$. This inflates type I error under $H_0^{(2)}$ for
  estimators with strong stage-1 instruments. The Sobel/delta test
  handles $H_0^{(2)}$ correctly via the delta-method SE. When
  $\alpha_M$ is expected to be nonzero and signals are not sparse,
  `"delta"` may therefore be preferable; `"composite"` is most
  beneficial when signals are sparse (the point-null $H_0^{(1)}$ is
  the dominant case).
- The JT-comp approximation is valid when $\text{Var}(a), \text{Var}(b) < 1.5$
  (approximately $n < 2000$ with sparse signals). ICONIC clamps variances
  to $[1, 1.5]$ to enforce this.
- Type I error is slightly inflated at small p-values (e.g., $\sim$15× at
  the genome-wide significance level $5 \times 10^{-7}$), as documented by
  Huang (2019) and Du et al. (2023). At conventional thresholds
  ($\alpha = 0.05$), the inflation is modest.

## Step 4: Stress-test — how robust is each estimator?

```r
# n_iter = 10 for fast vignette rendering; use 50+ in practice.
sens <- iconic_sensitivity(data, diagnosis = diag, n_iter = 10)
print(sens)
# Degradation surface: bias vs instrument exogeneity violations.
# Tipping points: first rho_G2 (at rho_G1 = 0) where |bias| > 0.10.
# Texture: auto-trained hybrid GAN + copula from data.
```

The surface sweeps a 2D grid of instrument-independence violations
($rho_{G1}$ x $rho_{G2}$), calibrated to your sample size and confounding
level. See the example walkthrough below for sample output.

### Auto-trained texture model

`iconic_sensitivity()` auto-trains a hybrid generative texture model from
your `iconic_data` object. The model has two independently trained
components: a sample-level torch GAN that learns the joint distribution of
per-sample exposure, outcome, mediator level, and encoded covariates, and a
feature-level Gaussian copula that learns the full joint distribution of the
mediator (M) panel — marginal distributions (empirical CDF by default, with
parametric fallback via KS test) and cross-feature dependence. The
`texture_source` field in the return object reports how the texture model was
obtained:

- `"auto-trained from data"` — trained on first call (default)
- `"data-attached GAN"` — reused from `iconic_data(trained_gan = ...)`
- `"user-supplied GAN"` — passed via the `trained_gan` argument

The `torch` package is required (it is a hard dependency of `iconic`).

To reuse a pre-trained model across multiple calls, attach it at
construction:

```r
gan <- train_gan_on_real_data(my_training_frame, epochs = 300)
data <- iconic_data(Z = my_Z, Y = my_Y, ..., trained_gan = gan)
# Now iconic_sensitivity(data) reuses the attached GAN without retraining.
```

### Data-calibrated confounding parameters

The `confounding` argument controls how the held-fixed confounding parameters
(`mo_confounding`, `omega_1`, `omega_2`) are set:

- `confounding = "default"` (default): fixed defaults (0.8, 0.7, 0.7).
- `confounding = "inferred"`: calls `infer_confounding()` to estimate
  parameters from the data. Parameters that cannot be inferred fall back to
  defaults with warnings.
- `confounding = "manual"`: use the explicitly supplied `mo_confounding`,
  `omega_1`, `omega_2` arguments.

```r
# Use data-inferred confounding parameters
sens <- iconic_sensitivity(data, diagnosis = diag,
                           confounding = "inferred", n_iter = 50)
print(sens$inferred_confounding)  # shows inferred values and warnings
```

See [Data-calibrated confounding
parameters](#data-calibrated-confounding-parameters) for details.

## Step 5: Recommend — which model should you trust?

```r
rec <- iconic_recommend(data, diagnosis = diag,
                        estimate = est, sensitivity = sens)
print(rec)
# Recommended estimator with tier, point estimates, and robustness score.
```

Estimators are ranked by tier (A > B > C > D) and, within tier, by robustness
(lower max |bias| across the violation surface = higher rank).

---

# Prospective analysis: what if you don't have instruments yet?

If you have exposure, mediator, and outcome data but no instruments or
negative controls, ICONIC can show you how your estimates would change if you
added them.

```r
data <- iconic_data(Z = my_exposure, Y = my_outcomes, M = my_mediator)
prospect <- iconic_prospect(data, n_iter = 50)
print(prospect)
# Phase 1: strength surface — how estimates converge as gamma_G increases.
# Phase 2: prospective simulation at target strength
#           if you collected an instrument of that quality.
# Reports F-statistics and recommends
# which estimator would be best.
# Texture: auto-trained from data .
```

`iconic_prospect()` also auto-trains a hybrid texture model (GAN + copula)
from your data and supports the `confounding` argument. In the prospective
setting (no instruments or NCs), most confounding parameters cannot be
inferred and will fall back to defaults with warnings — this is an honest
limitation, not a silent failure.

This is useful for study design: before spending resources on genotyping or
collecting negative controls, you can see whether the investment would change
your conclusions.

---

# Data-calibrated confounding parameters

`infer_confounding()` estimates the held-fixed confounding parameters that
`iconic_sensitivity()` and `iconic_prospect()` use, so the degradation
surface is calibrated to your data rather than using fixed defaults.

## What it infers

| Parameter | Method | Requires | Assumption |
|-----------|--------|----------|------------|
| `conf_strength` ($\delta$) | UNADJ–IV2SLS gap | valid G ($F \ge 10$) | exclusion |
| `mo_confounding` ($\delta_{mo}$) | IV2SLS–IV2SLS2 NIE gap | valid G + Gm | exclusion |
| `omega_1`, `omega_2` | $\sqrt{R^2}$ of W on resid. Y | W (neg. controls) | valid NCs |
| `k` | parallel analysis (Horn, 1965) | $\ge 5$ features | factor model |

$\rho_{G1}$ and $\rho_{G2}$ (instrument–confounder correlations) are
unestimable because $U$ is unobserved. These always remain sweep variables in
the degradation surface.

The inference uses estimator validity to calibrate a benchmark whose purpose
is to test estimator validity — this is circular. The inferred values should
be interpreted as best-case calibrations under the stated assumptions, not as
ground truth. The `confounding = "default"` mode avoids this circularity
entirely and remains the recommended starting point.

## Usage

```r
# Standalone: inspect inferred parameters before using them
diag <- iconic_diagnose(data)
est  <- iconic_estimate(data, diagnosis = diag)
conf <- infer_confounding(data, diagnosis = diag, estimate = est)
print(conf)
# <iconic_confounding>
#   conf_strength     0.041  (UNADJ-IV2SLS gap (NDE))
#     warning: inferred delta is very low; ...
#   mo_confounding    0.077  (IV2SLS-IV2SLS2 NIE gap)
#   omega_1           0.073  (sqrt(R²) of W on Y residualized on Z+C)
#     warning: composite: coverage × confounder strength, not pure coverage
#   omega_2           0.749  (sqrt(R²) of W on Y residualized on Z+C)
#     warning: composite: coverage × confounder strength, not pure coverage
#   k                 1 [CI: 1, 2]  (parallel analysis (Horn, 1965))
#   Unavailable: rho_G1, rho_G2

# Integrated: pass confounding = "inferred" to iconic_sensitivity()
sens <- iconic_sensitivity(data, diagnosis = diag,
                           confounding = "inferred", n_iter = 50)
# sens$inferred_confounding contains the same object
# sens$omega_1, sens$omega_2 reflect the inferred values
```

Each parameter slot reports its `estimate`, `method`, `assumption`,
`available` (logical), and `warning` (character or NULL). Parameters that
cannot be inferred (missing instruments, NCs, or insufficient features) are
marked `available = FALSE` and fall back to defaults.

---

# Example walkthrough: the full workflow on simulated data

This section demonstrates the complete workflow on a small simulated dataset,
with real output so you can see what each step produces. The parameters are
kept small for fast execution; in practice you would use `n_iter = 30–50`
and larger grids.

## Step 1: Create an iconic_data object

We generate a mediation dataset with M-O confounding, two instruments (G, Gm),
and path-specific negative controls (W1, W2):

```{r walkthrough-step1}
library(iconic)
set.seed(42)

dat <- generate_toy_data(n = 200, n_features = 10, mo_confounding = 0.8,
                         phi = 0.8, separate_U = TRUE,
                         omega_1 = 0.7, omega_2 = 0.7, seed = 42)
idata <- iconic_data(Z = dat$Z, Y = dat$Y, M = dat$M,
                     G = dat$G[, 1], Gm = dat$Gm,
                     W1 = dat$W1, W2 = dat$W2)
print(idata)
```

The data object reports 200 samples, 10 outcome features, 1 mediator, and all
identifying resources present (G, Gm, W, W1/W2).

## Step 2: Diagnose — which estimators are eligible?

```{r walkthrough-step2}
diag <- iconic_diagnose(idata)
print(diag)
```

Both instruments are strong (partial F = 174.1 for G, 1725.7 for Gm). The A2
and A2' screens pass (all 10 controls valid), and completeness is satisfied
(10 valid NCs vs k=1). All 8 estimators are eligible.

The full instrument-strength distributions are available at
`diag$instrument_strength$F_G` and `diag$instrument_strength$F_Gm` (numeric
vectors). With a single mediator here, each is length 1; with a mediator
panel (see below), they contain one entry per transcript.

### Panel-level gating with heterogeneous instrument quality

When the mediator panel has many transcripts, each with its own
mediator-instrument (e.g. one cis-eQTL per isoform), instrument strength
varies across the panel. The scalar median can hide a large fraction of
weak instruments. The `gm_threshold = list(E, R)` argument gates eligibility
on the distribution: a Gm-dependent method is eligible if at least fraction
`E` of transcripts have partial F >= `R`.

```{r walkthrough-step2b}
# Build a multi-mediator panel where eQTL strength varies
set.seed(7)
n <- 200; nm <- 50
G <- rnorm(n)
Z <- 0.3 * G + rnorm(n, 0, 0.9)
# Strong eQTLs for the first 25 mediators, weak for the rest
Gm_mat <- matrix(NA, nm, n)
M_mat  <- matrix(NA, nm, n)
for (m in 1:nm) {
  gm_m <- if (m <= 25) rnorm(n) else rnorm(n, 0, 0.3)
  Gm_mat[m, ] <- gm_m
  M_mat[m, ]  <- 0.5 * gm_m + 0.2 * Z + rnorm(n, 0, 0.7)
}
Y <- 0.1 * Z + 0.3 * M_mat[1, ] + rnorm(n, 0, 0.8)
W  <- matrix(rnorm(10 * n), 10, n)

idata2 <- iconic_data(Z = Z, Y = matrix(Y, 1), M = M_mat,
                      G = G, Gm = Gm_mat, W = W)

# Legacy scalar rule: median F_Gm vs min_f (default 10)
diag_scalar <- iconic_diagnose(idata2)
cat("Scalar rule:\n")
print(diag_scalar)

# Panel rule: eligible if >= 50% of transcripts have F_Gm >= 10
diag_panel <- iconic_diagnose(idata2,
  gm_threshold = list(E = 0.5, R = 10))
cat("\nPanel rule (E=50%, R=10):\n")
print(diag_panel)

# Inspect the full distribution
fgm <- diag_panel$instrument_strength$F_Gm
cat("\nF_Gm distribution (", length(fgm), " mediators):\n", sep = "")
print(summary(fgm))
for (R in c(3, 5, 10))
  cat(sprintf("  F_Gm >= %2d: %5.1f%%\n", R,
              100 * mean(fgm >= R, na.rm = TRUE)))
```

Under the scalar rule, the median F_Gm is high (the strong half dominates the
median), so IV2SLS2 is eligible — but the 25 weak-eQTL transcripts will
produce `NA` estimates at the per-transcript gate. Under the panel rule with
E=50%, R=10, only ~50% of transcripts clear R=10, so the method is on the
eligibility boundary. Lowering R to 5 or E to 0.4 would make it clearly
eligible; the choice depends on how much weak-instrument bias you're willing
to tolerate. The distribution summary lets you pick E and R informed by the
actual spread of eQTL strengths.

## Step 3: Estimate — fit all eligible estimators

```{r walkthrough-step3}
est <- iconic_estimate(idata, diagnosis = diag)
head(est, 8)
cat("True NDE =", dat$true_NDE, "  True NIE =", dat$true_NIE, "\n")
```

The true NDE = 0.10 and NIE = 0.15. IV2SLS2 (NDE = 0.123) and PGC2Gm (NDE =
0.107) are closest to the true NDE, while UNADJ (NDE = 0.067) and COCA (NDE =
-0.226) are substantially biased — as expected under M-O confounding.

Two optional arguments control which transcripts get estimates:

- `min_f` (default 10): the per-transcript weak-instrument gate. Transcripts
  whose instrument F falls below this value receive `NA` for that method.
  Lower it (e.g. `min_f = 3`) to keep weak-instrument transcripts at the cost
  of potential weak-IV bias.
- `run_all` (default FALSE): when TRUE, overrides diagnosis eligibility and
  runs every method whose required data exists. The per-transcript `min_f`
  gate still applies.

```r
# Exploratory: run all methods, keep transcripts with F >= 3
est_all <- iconic_estimate(idata2, run_all = TRUE, min_f = 3)
```

## Step 4: Infer confounding parameters

```{r walkthrough-step4}
conf <- infer_confounding(idata, diagnosis = diag, estimate = est)
print(conf)
```

The inferred `conf_strength` is low (0.041) because the strong instrument (F =
174) already removes most of the confounding bias that UNADJ carries — the gap
between UNADJ and IV2SLS is small. The `omega_2` estimate (0.749) is close to
the true coverage (0.7), while `omega_1` is low (0.073) because W1's
correlation with the residualized outcome is weak in this particular draw. The
`k` estimate (1, CI: [1, 2]) correctly identifies a single latent confounder.
The warnings flag the circularity and composite-measure caveats.

## Step 5: Sensitivity analysis with auto-trained GAN

```{r walkthrough-step5}
sens <- iconic_sensitivity(idata, diagnosis = diag,
                           rho_G1_grid = c(0, 0.3),
                           rho_G2_grid = c(0, 0.3),
                           n_iter = 5, n_features = 3,
                           gan_epochs = 20)
print(sens)
```

The texture model was auto-trained from the data (sample-level GAN +
feature-level Gaussian copula). The degradation surface shows that IV2SLS2
and PGC2Gm are robust across the grid, while UNADJ and COCA tip immediately.
The specific bias values depend on the learned texture; with the full
manuscript settings (25 replicates, 5×5 grid), IV2SLS2 has max |NDE bias| =
0.090 and PGC2Gm has max |NDE bias| = 0.124, so IV2SLS2 is the more robust
estimator under the learned texture — the reverse of the toy-simulation
benchmark (Figure 3), where PGC2Gm is more robust.

### With inferred confounding parameters

```{r walkthrough-step5b}
sens_inf <- iconic_sensitivity(idata, diagnosis = diag,
                               rho_G1_grid = c(0, 0.3),
                               rho_G2_grid = c(0, 0.3),
                               n_iter = 5, n_features = 3,
                               confounding = "inferred", gan_epochs = 20)
print(sens_inf)
print(sens_inf$inferred_confounding)
```

When confounding parameters are inferred, the surface uses the data-calibrated
values (mo_confounding = 0.077, omega_1 = 0.073, omega_2 = 0.749) instead of
defaults (0.8, 0.7, 0.7). The lower mo_confounding reduces the bias of
single-instrument methods, while the asymmetric omega values shift the
degradation pattern. The `inferred_confounding` object is attached to the
return value for transparency.

## Step 6: Recommend — which estimator should you trust?

```{r walkthrough-step6}
rec <- iconic_recommend(idata, diagnosis = diag, estimate = est,
                        sensitivity = sens)
print(rec)
```

The recommended estimator is the one with the lowest worst-case bias across
the degradation surface among the highest-tier eligible estimators. With the
full manuscript settings, IV2SLS2 is recommended (tier A, max |bias| = 0.090)
over PGC2Gm (tier A, max |bias| = 0.124). Both are identified estimators
using two instruments plus path-specific NCs. The ranking reflects IV2SLS2's
superior robustness under the learned data texture. Note that this ordering
is the reverse of the toy-simulation benchmark (Figure 3), where PGC2Gm is
the more robust estimator — the learned texture shifts the relative
robustness of the two identified estimators.

## Step 7: Prospective analysis (for data without instruments)

For users who have Z, M, Y but no instruments or NCs, `iconic_prospect()`
simulates what estimates they could expect if they collected such data:

```{r walkthrough-step7}
bare <- iconic_data(Z = rnorm(200), Y = matrix(rnorm(200 * 5), 5, 200),
                    M = rnorm(200))
prospect <- iconic_prospect(bare, gamma_G_grid = c(0.3, 0.6),
                            n_iter = 5, n_features = 3, gan_epochs = 20)
print(prospect)
```

Phase 1 shows UNADJ bias decreasing as the instrument strengthens, while
IV2SLS2 and PGC2Gm remain near zero. Phase 2 reports prospective estimates at
the target strength: with the full manuscript settings (25 replicates,
$\gamma_G = 0.6$), the first-stage partial $F$-statistics are 264 for the
exposure instrument and 136 for the mediator instrument, both well above the
$F \geq 10$ threshold. The recommended estimator if instruments are collected
is IV2SLS2.

---

# Simulation and benchmarking functions

The simulation functions are exposed for methodologists who want full control
over the data-generating process and benchmarking. Most users will interact
with `iconic_diagnose()`, `iconic_estimate()`, `iconic_sensitivity()`, and
`iconic_recommend()` directly (see the sections above).

## Running the mediation simulation

```{r mediation-sim}
# Generate one dataset with M-O confounding
dat <- generate_toy_data(n = 300, n_features = 5,
                                  mo_confounding = 0.8, seed = 42)
cat("True NDE =", dat$true_NDE, "  True NIE =", dat$true_NIE, "\n")

# Run all eligible mediation estimators
res <- analyze_mediation_robust(dat)
aggregate(NDE ~ method, res, function(v) round(mean(v, na.rm = TRUE), 3))
aggregate(NIE ~ method, res, function(v) round(mean(v, na.rm = TRUE), 3))
```

For a proper benchmark with many replicates:

```{r mediation-bench, eval=FALSE}
# 100 replicates, 20 features, M-O confounding = 0.8
bench <- run_mediation_sim(n_iter = 100, n_features = 20,
                           mo_confounding = 0.8, n_cores = 4)
bench$summary[, c("method", "NDE_bias", "NIE_bias",
                     "NDE_rmse", "NIE_rmse")]
```

## Sweeping mediation parameters

```{r mediation-sweep, eval=FALSE}
# Sweep confounding strength
sweep_conf <- sweep_mediation_param("conf_str", c(0.2, 0.4, 0.6, 0.8, 1.0),
                                    n_iter = 50, mo_confounding = 0.8)

# Sweep proxy quality
sweep_omega <- sweep_mediation_param("w_signal",
                                     c(0.2, 0.4, 0.6, 0.7, 0.8, 0.9),
                                     n_iter = 50, mo_confounding = 0.8)

# Sweep sample size (bias should NOT shrink — it's structural)
sweep_n <- sweep_mediation_param("n_samples", c(100, 200, 500, 1000),
                                 n_iter = 50, mo_confounding = 0.8)

# Null Type I error across confounding strength
null_t1e <- sweep_mediation_null_by_conf(
    c(0.2, 0.4, 0.6, 0.8, 1.0), n_iter = 50)
```

## Two-stage MR mediation

When a mediator-specific instrument $G_m$ is available (`phi > 0`), the
2-stage MR estimator (IV2SLS2) instruments both $Z$ and $M$:

```{r iv2sls2-demo}
# Generate one dataset with M-O confounding AND a mediator instrument
dat <- generate_toy_data(n = 500, n_features = 5,
                                  mo_confounding = 0.8,
                                  phi = 0.8, seed = 42)
cat("True NDE =", dat$true_NDE, "  True NIE =", dat$true_NIE, "\n")
cat("Gm present:", !is.null(dat$Gm), "\n")

# Run all eligible mediation estimators (IV2SLS2 included because Gm is present)
res <- analyze_mediation_robust(dat)
aggregate(NDE ~ method, res, function(v) round(mean(v, na.rm = TRUE), 4))
aggregate(NIE ~ method, res, function(v) round(mean(v, na.rm = TRUE), 4))
```

IV2SLS2 can also be called directly:

```{r iv2sls2-direct, eval=FALSE}
# Direct usage: fit_iv2sls_mediation2() with both instruments
# w can be a vector (single NC) or a matrix (full NC panel)
fit <- fit_iv2sls_mediation2(
  y = dat$Y[, 1], Z = dat$Z, M = dat$M,
  g = dat$G[, 1], gm = dat$Gm, w = dat$W
)
fit$NDE   # natural direct effect
fit$NIE   # natural indirect effect
```

```{r iv2sls2-bench, eval=FALSE}
# 50 replicates with mediator instrument
bench <- run_mediation_sim(n_iter = 50, n_features = 10, n_samples = 500,
                           mo_confounding = 0.8, phi = 0.8, n_cores = 4)
bench$summary[, c("method", "NDE_bias", "NIE_bias",
                     "NDE_rmse", "NIE_rmse")]
```

## Sweeping mediator-instrument strength

The `phi` parameter controls the strength of the mediator instrument
$G_m \to M$. Sweeping $\varphi$ from 0 to 1.0 shows IV2SLS2 bias converging
to zero as the instrument strengthens:

```{r phi-sweep, eval=FALSE}
# Sweep mediator-instrument strength
sweep_phi <- sweep_mediation_param("phi",
                                   c(0, 0.2, 0.4, 0.6, 0.8, 1.0),
                                   n_iter = 30, n_features = 10,
                                   n_samples = 500,
                                   mo_confounding = 0.8, n_cores = 4)
# IV2SLS2 bias vs phi
iv2sls2 <- sweep_phi$summary[sweep_phi$summary$method == "IV2SLS2", ]
iv2sls2[, c("param_value", "NDE_bias", "NIE_bias", "NDE_rmse", "NIE_rmse")]
```

## Mediator-instrument independence check

`nc_independence_check_gm()` tests whether the mediator instrument $G_m$ is
independent of the negative controls (A2' screen):

```{r gm-independence}
# Generate a dataset with a mediator instrument
dat <- generate_toy_data(n = 500, n_features = 10,
                                  mo_confounding = 0.8,
                                  phi = 0.8, seed = 1)

# A2': Are any controls associated with the mediator instrument?
gm_check <- nc_independence_check_gm(dat)
head(gm_check)
table(gm_check$verdict)
```

On simulation data, $G_m \perp W$ holds by construction, so all controls
return "valid". On real data, controls that survive FDR correction should be
dropped before running IV2SLS2 — they may carry eQTL or allele-specific
effects that violate the mediator-instrument-independence assumption.

When `dat$Gm` is absent (i.e. `phi = 0`), the function returns `NULL`:

```{r gm-independence-absent, eval=FALSE}
dat_no_gm <- generate_toy_data(n = 500, n_features = 10, seed = 1)
nc_independence_check_gm(dat_no_gm)  # NULL: no mediator instrument
```

## Imperfect instrument independence

`generate_toy_data()` supports parameters that violate instrument
independence, modelling realistic genomic structure:

| Parameter | Meaning | Default |
|-----------|---------|---------|
| `rho_G1` | Correlation of $G_1$ (exposure instrument) with $U_{XM}$ | 0 |
| `rho_G2` | Correlation of $G_2$ (mediator instrument) with $U_{MY}$ | 0 |
| `rho_pop` | Shared population structure $P$ inducing $G_1$–$G_2$ correlation | 0 |
| `separate_U` | Draw separate confounders $U_{XM}$ and $U_{MY}$ for each path | FALSE |
| `omega_1` | Coverage of $U_{XM}$ by path-specific NCs $W_1$ | NULL (= `w_signal`) |
| `omega_2` | Coverage of $U_{MY}$ by path-specific NCs $W_2$ | NULL (= `w_signal`) |

When any parameter is non-default, the output additionally includes
`G1`, `G2`, `W1`, `W2`, `U_XM`, `U_MY`, and (when `rho_pop > 0`) `P`.

## Demonstrating PGC-2

`fit_pgc_mediation2()` uses path-specific negative controls to purge
confounding at both the exposure and mediator stages. Unlike IV2SLS2, it does
not require instrument exogeneity — the bridge absorbs the confounder
regardless of instrument violation.

```{r pgc2-demo}
# Generate a dataset with imperfect independence, separate confounders
dat <- generate_toy_data(n = 500, n_features = 5,
                                  mo_confounding = 0.8, phi = 0.8,
                                  rho_G1 = 0.3, rho_G2 = 0.3,
                                  separate_U = TRUE,
                                  omega_1 = 0.7, omega_2 = 0.7, seed = 42)
cat("True NDE =", dat$true_NDE, "  True NIE =", dat$true_NIE, "\n")
cat("W1 present:", !is.null(dat$W1),
    "  W2 present:", !is.null(dat$W2), "\n")

# Run all eight estimators
# (PGC2, PGC2Gm included because W1/W2 are present)
res <- analyze_mediation_robust(dat)
aggregate(NDE ~ method, res, function(v) round(mean(v, na.rm = TRUE), 4))
aggregate(NIE ~ method, res, function(v) round(mean(v, na.rm = TRUE), 4))
```

PGC-2 can also be called directly:

```{r pgc2-direct}
# Without mediator instrument (pure NC identification)
fit_nc <- fit_pgc_mediation2(dat$Y[, 1], dat$Z, dat$M, dat$G1,
                             dat$W1, dat$W2)
cat("PGC2  NDE =", round(fit_nc$NDE, 4),
    " NIE =", round(fit_nc$NIE, 4), "\n")

# With (possibly imperfect) mediator instrument
fit_gm <- fit_pgc_mediation2(dat$Y[, 1], dat$Z, dat$M, dat$G1,
                             dat$W1, dat$W2, gm = dat$Gm)
cat("PGC2Gm NDE =", round(fit_gm$NDE, 4),
    " NIE =", round(fit_gm$NIE, 4), "\n")
```

## Sweeping the degradation surface

The degradation surface sweeps both instrument exogeneity violations
simultaneously ($\rho_{G1}$ and $\rho_{G2}$):

```{r degradation-surface, eval=FALSE}
# 2D sweep: both instrument exogeneity violations
results <- data.frame()
for (rg1 in c(0, 0.1, 0.2, 0.3, 0.5)) {
  for (rg2 in c(0, 0.1, 0.2, 0.3, 0.5)) {
    sweep_res <- sweep_mediation_param(
      "rho_G2", rg2,
      n_iter = 50, n_samples = 500, n_features = 10,
      mo_confounding = 0.8, phi = 0.8, rho_G1 = rg1,
      separate_U = TRUE, omega_1 = 0.7, omega_2 = 0.7)
    smry <- sweep_res$summary
    for (m in c("IV2SLS2", "PGC2Gm")) {
      row <- smry[smry$method == m, ]
      results <- rbind(results, data.frame(
        rho_G1 = rg1, rho_G2 = rg2, method = m,
        NDE_bias = row$NDE_bias, NIE_bias = row$NIE_bias))
    }
  }
}
```

---

# The generative pipeline

## Training a texture model

The texture model is a hybrid of two independently trained components,
trained by `train_gan_on_real_data()`:

```{r train}
# 1. Load real data as features x samples matrices + a covariate frame.
#    Here we use the built-in example; swap in your own:
#    load_real_input_data(Z_matrix = my_gdm, Y_matrix = my_expr,
#                         covariates_df = my_meta)
input <- load_real_input_data(example = TRUE)
str(input, max.level = 1)

# 2. Train the generator (requires torch).
#    epochs = 20 for fast vignette rendering; use 100-300 in practice.
gan <- train_gan_on_real_data(input$gan_training_data,
                                 epochs = 20, seed = 1)
gan
```

The model has two components:

- **Sample-level GAN** (`torch`): an MLP generator and discriminator trained
  with a non-saturating BCE objective, Adam ($\beta_1 = 0.5$), one-sided
  label smoothing, and per-column normalization. This GAN learns the joint
  distribution of per-sample exposure, outcome, mediator level, and encoded
  covariates (one row per sample).
- **Feature-level Gaussian copula** (`train_feature_texture()`): learns the
  full joint distribution of the mediator (M) panel. Each feature's marginal
  is fitted as an empirical CDF by default, with a parametric fallback
  (normal, log-normal, gamma, or beta, selected by AIC) used when the
  parametric fit passes a Kolmogorov-Smirnov test at $p > 0.05$. The user may
  override via the `marginal_method` parameter (`"auto"`, `"empirical"`, or
  `"parametric"`). Cross-feature dependence is captured by the Gaussian
  copula correlation matrix of the normal-transformed scores. During
  simulation, copula draws replace the parametric Gaussian noise in the
  mediator structural equations, preserving realistic marginal shapes and
  correlations from the user's cohort while maintaining the closed-form
  ground truth.

The `torch` package is a hard dependency (listed in `Imports`).
`train_gan_on_real_data()` errors if `torch` is unavailable.

We can inspect fidelity — real vs. synthetic marginals — before trusting the
data:

```{r diag, fig.height=5}
plot_gan_diagnostics(gan, input$gan_training_data)
```

### Auto-training in the model selection workflow

`iconic_sensitivity()` and `iconic_prospect()` automatically train a hybrid
texture model (sample-level GAN + feature-level copula) from the user's
`iconic_data` object when no `trained_gan` is supplied. The auto-training
converts the `iconic_data` object to the `load_real_input_data()` format and
calls `train_gan_on_real_data()` with a lighter epoch budget (default 100,
vs. 300 for standalone training). The feature-level copula is trained from
the mediator matrix via `train_feature_texture()`.

Priority resolution: explicit `trained_gan` argument > `data$trained_gan`
(attached at `iconic_data()` construction) > auto-train from `data`.

---

# Generalization 1: multiple confounders

`run_single_iteration()` carries a latent matrix
$U \in \mathbb{R}^{n\times k}$ (`n_confounders = k`), with per-pathway
multipliers (`MMExp`, `MMOut`, `MMCon`, `MMCpG`) scaling how strongly the
confounders load into the exposure, outcome, controls, and methylation
respectively.

```{r iterate}
dat <- run_single_iteration(
  gan,
  n_synthetic_samples = 500,
  n_features          = 12,
  n_confounders       = 2,       # two latent confounders
  beta_Z = 0.10, alpha_M = 0.50, beta_M = 0.30,
  conf_strength       = 0.8,
  seed                = 42
)
cat("True total effect tau =", dat$true_total, "\n")
# 0.10 + 0.50*0.30 = 0.25

# Output has the shape run_methods()/
# analyze_methods_robust() expect.
res <- analyze_methods_robust(dat)
aggregate(beta ~ method, res, function(v) round(mean(v, na.rm = TRUE), 3))
```

To add mediator-outcome confounding, set `mo_confounding`:

```{r iterate-mo}
dat_mo <- run_single_iteration(
  gan,
  n_synthetic_samples = 500,
  n_features          = 12,
  n_confounders       = 1,
  mo_confounding      = 0.8,     # U1 -> M (mediator-outcome confounding)
  seed                = 42
)
cat("True NDE =", dat_mo$true_NDE, "  True NIE =", dat_mo$true_NIE, "\n")

# Run mediation estimators
res_mo <- analyze_mediation_robust(dat_mo)
aggregate(NDE ~ method, res_mo,
           function(v) round(mean(v, na.rm = TRUE), 3))
aggregate(NIE ~ method, res_mo,
           function(v) round(mean(v, na.rm = TRUE), 3))
```

To add horizontal pleiotropy (a direct $G \to Y$ path), set `pleio`:

```{r iterate-pleio}
dat_pl <- run_single_iteration(
  gan,
  n_synthetic_samples = 500,
  n_features          = 12,
  n_confounders       = 1,
  pleio = 0.10,  # direct G -> Y path
                # (violates exclusion restriction)
  seed                = 42
)
# IV2SLS is now biased because G affects Y directly, not only through Z
res_pl <- analyze_methods_robust(dat_pl)
aggregate(beta ~ method, res_pl,
           function(v) round(mean(v, na.rm = TRUE), 3))
```

---

# Generalization 2: pluggable negative controls

A negative-control model is any function with the signature

```r
function(U, covariates, params) -> W   # returns an n x n_features matrix
```

so you are never locked into one theory of where the controls come from.
Built-ins:

- **`nc_proxy`** — each control is a coverage-weighted mixture of the
  captured confounders plus noise.
- **`nc_cpg`** — simulates spatially correlated CpG methylation (an AR(1)
  process along the genome) partly driven by the confounders, then forms
  each control as a sparse linear prediction from the methylation
  ("CpG-predicted expression").

```{r nc}
list_nc_models()

# Use the CpG-predicted-expression controls instead of the direct proxy:
dat_cpg <- run_single_iteration(gan, n_features = 12, n_confounders = 1,
                                nc_model = "cpg", coverage = 0.7, seed = 7)
dim(dat_cpg$W)

# Or supply your own mechanism — anything honouring the contract:
my_nc <- function(U, covariates, params) {
  # controls = first confounder + independent noise
  scale(matrix(U[, 1], nrow(U), params$n_features) +
        matrix(rnorm(nrow(U) * params$n_features), nrow(U)))
}
dat_custom <- run_single_iteration(gan, n_features = 8,
                                      nc_model = my_nc, seed = 8)
dim(dat_custom$W)
```

The `coverage`/`captured` knobs control whether the negative controls span
the same confounder subspace that drives the outcome. Setting `captured` to a
strict subset models controls that miss part of $U$ — the failure mode a
sensitivity analysis should probe.

---

# The sensitivity analysis

## Sweeping confounding scenarios

`gan_sensitivity()` benchmarks every estimator across a grid of confounding
strength, negative-control coverage, and number of latent confounders:

```{r sens}
# n_iter = 5 for fast vignette rendering; use 25+ in practice.
sens <- gan_sensitivity(
  gan,
  conf_grid     = c(0.2, 0.5, 0.8),
  coverage_grid = c(0.3, 0.6, 1.0),
  k_grid        = c(1, 2),
  nc_model      = "proxy",
  n_iter        = 5,
  n_samples     = 400,
  n_features    = 12
)
head(sens$summary, 8)
```

A heatmap reads off where a method's error blows up:

```{r heat}
plot_sensitivity_heatmap(sens, metric = "rmse", method = "IV2SLS", k = 1)
```

## Which estimator is preferred?

`recommend_estimator()` ranks methods by RMSE within each scenario and
identifies an overall pick with the smallest worst-case RMSE across scenarios
(a minimax choice; UNADJ is excluded as a reference).

```{r rec}
rec <- recommend_estimator(sens)
rec$overall
rec$worst_case
```

## Do the negative controls hold?

`nc_validity_check()` focuses on the negative-control estimators (COCA, PGC)
as coverage drops and confounders multiply. It enforces the identifiability
condition: the bridge-function estimators are point identified only when
there are at least as many valid negative controls as confounders.

```{r validity}
# n_iter = 5 for fast vignette rendering; use 20+ in practice.
validity <- nc_validity_check(
  gan,
  coverage_grid    = c(0.2, 0.5, 1.0),
  k_grid           = c(1, 2, 3),
  conf_strength    = 0.8,
  n_valid_controls = 1,  # your design provides
                          # 1 distinct valid control
  n_iter           = 5,
  n_features       = 12
)
validity$verdict
```

Rows with `k > n_valid_controls` are flagged under-identified: no estimator
built on those controls can recover $\tau$ there, regardless of sample size.

## Empirical negative-control validity diagnostics

On real data, the two identifying assumptions that negative controls must
satisfy can be tested directly:

1. $W \perp Z \mid C$ — controls are not affected by the exposure (A1).
2. $W \perp G \mid C$ — controls are independent of the instrument (A2).

`iconic` provides three exported diagnostic functions:

```{r nc-diagnostics}
# Generate a dataset to illustrate (on real data, pass your fitted dataset)
dat <- run_single_iteration(gan, n_features = 10,
                               n_confounders = 1, seed = 1)

# A1: Are any controls associated with the exposure after adjusting for C?
screen_Z <- nc_validity_screen(dat)
head(screen_Z)
table(screen_Z$verdict)

# A2: Are any controls associated with the
# instrument after adjusting for C?
screen_G <- nc_independence_check(dat)
head(screen_G)
table(screen_G$verdict)

# Completeness: enough valid controls for the confounder dimensionality?
comp <- nc_completeness_check(dat)
comp$completeness
comp$n_valid_controls
comp$k
```

- `nc_validity_screen()` regresses each $W_f$ on $Z + C$ and applies
  Benjamini-Hochberg FDR control at 0.10. Controls surviving FDR are flagged
  "drop: associated with Z" — they violate A1.
- `nc_independence_check()` computes the partial correlation of each $W_f$
  with $G$ after residualising both on $C$, and applies BH-FDR. Controls
  surviving FDR are flagged "drop: associated with G" — they violate A2.
- `nc_completeness_check()` counts controls that pass both screens and
  compares to $k$.

> **Interpretation on simulation data.** In the simulation, $W$ shares the
> confounder $U$ with $Z$, so $W$ is associated with $Z$ through $U$ — the A1
> screen will flag controls as "associated with Z." This is expected: the
> screen detects the confounder-driven association, which is precisely the
> association the NC estimators exploit. The screen is designed for real
> data, where a control being associated with $Z$ after adjusting for $C$
> may indicate a direct $Z \to W$ path (a true violation) rather than a
> shared confounder.

## Pleiotropy sensitivity sweep

`gan_pleiotropy_sensitivity()` benchmarks every estimator across a grid of
horizontal-pleiotropy strengths and confounding strengths, running both an
alternative arm (true effect $= \tau$) and a null arm (true effect $= 0$) so
that bias and Type I error can be read off the same sweep.

```{r pleio-sens}
# n_iter = 5 for fast vignette rendering; use 25+ in practice.
pleio_sens <- gan_pleiotropy_sensitivity(
  gan,
  pleio_grid  = c(0, 0.05, 0.10),   # direct G -> Y path strengths
  conf_grid   = c(0.2, 0.5, 0.8),
  tau = 0.25,  # true total effect
                # for the alternative arm
  n_iter      = 5,
  n_features  = 10,
  n_samples   = 500
)
head(pleio_sens$summary, 12)
```

The `pleio` parameter is available in both `run_single_iteration()` and
`generate_toy_data()`, so custom sweeps can be built on either generator.

## Instrument strength and the weak-IV guard

The IV/2SLS estimator is consistent only when the instrument is strong — the
first-stage partial $F$ for the excluded instrument should be $\ge 10$. The
package uses the partial F (squared $t$-statistic for $G$ conditional on $W$
and covariates) rather than the overall model $F$.

`iconic_diagnose()` computes the partial F for every instrument and every
mediator in the panel. The full distributions are stored at
`diag$instrument_strength$F_G` and `diag$instrument_strength$F_Gm`. Two
gating layers control whether an instrument-dependent method runs:

1. **Method eligibility** (`iconic_diagnose()`): by default, a method is
   eligible if the median partial F meets `min_f` (default 10). For
   heterogeneous panels, supply `g_threshold`/`gm_threshold = list(E, R)` to
   gate on the distribution: eligible if at least fraction `E` of
   transcripts have F >= `R`.
2. **Per-transcript execution** (`iconic_estimate()` via `min_f`):
   transcripts whose instrument F falls below `min_f` receive `NA` for that
   method, preventing weak-IV-biased estimates from individual transcripts.
   `run_all = TRUE` overrides layer 1 (runs all data-present methods) but
   layer 2 still applies.

```{r iv-strength, eval=FALSE}
# Demonstrate the weak-IV guard with an irrelevant instrument
dat_weak <- run_single_iteration(NULL, n_synthetic_samples = 200,
    n_features = 5, effect_size = 0.25, conf_strength = 0.8, seed = 1)
dat_weak$G <- matrix(rnorm(1000), 200, 5)   # replace G with pure noise
res <- analyze_methods_robust(dat_weak)
# IV2SLS returns NA for all features (partial F < 10)
res[res$method == "IV2SLS", ]
```

The guard is applied in both `fit_iv2sls()` and `fit_iv2sls_mediation()`.

## Mediation sensitivity sweep

`gan_mediation_sensitivity()` extends the grid sweep to the mediation
setting, benchmarking NDE/NIE bias and Type I error across confounding
scenarios with mediator-outcome confounding:

```{r med-sens, eval=FALSE}
med_sens <- gan_mediation_sensitivity(
  gan,
  conf_grid      = c(0.2, 0.5, 0.8),
  coverage_grid  = c(0.3, 0.7, 1.0),
  k_grid         = 1,
  mo_confounding = 0.8,
  n_iter         = 50,
  n_features     = 12
)
head(med_sens$summary[, c("conf_strength", "coverage", "method",
                          "NDE_bias", "NIE_bias", "NIE_type1")])
```

---

# Feature-level correlations

Real omics data exhibit cross-feature correlations — e.g., placental
transcript expression forms co-expression modules with high within-module
and low between-module correlation. ICONIC supports this structure in two
places: a `feat_cor` parameter in the toy simulation, and feature-level
texture learning in the generative pipeline (Gaussian copula for the
mediator panel, residual correlation matrices for Y and W).

## Block-diagonal co-expression in the toy simulation

`generate_toy_data()` and `sweep_mediation_param()` accept a `feat_cor`
parameter that injects block-diagonal correlated noise into the outcome (Y)
and negative-control (W) panels. Features are divided into
`ceiling(sqrt(p))` equal-sized modules; within-module pairwise correlation
equals `feat_cor`, between-module correlation is 0. For `p = 10`: 4 modules
of sizes 3, 3, 2, 2. The mediator M is scalar by design, so `feat_cor` does
not apply to M.

```{r feat-cor-sweep, eval=FALSE}
# Sweep feature correlation strength
# rho_G1 = 0 ensures the NC bridge is live (A2 passes, all 8 estimators eligible)
fc_sweep <- sweep_mediation_param("feat_cor", c(0, 0.2, 0.4, 0.6, 0.8),
    n_iter = 50, mo_confounding = 0.8, phi = 0.8,
    rho_G1 = 0, separate_U = TRUE, omega_1 = 0.7, omega_2 = 0.7)
fc_sweep$summary

# Plot the sweep (3 panels: bias, RMSE, Type I error)
plot_feature_correlation_sweep(fc_sweep)
```

Empirically, estimator performance is invariant to `feat_cor`: NDE and NIE
bias change by less than 0.008 across the full sweep for all estimators
except PGC2 (which is inherently unstable due to bridge matrix inversion).
NIE Type I error is perfectly controlled (IV2SLS2 and PGC2Gm both at 0.00
at all `feat_cor` levels). The A2 screen passes at 89--96% of NC features,
all 8 estimators are eligible, and the tier ranking (IV2SLS2 > PGC2Gm >
PGC > COCA > PGC2 > IV2SLS > DIRECT > UNADJ) is identical at all
`feat_cor` levels. This invariance holds because correlated noise is
additive — it enters Y and W as multivariate normal noise on top of the
structural signal, without perturbing the coefficients that define the
ground truth (verified by an oracle regression test in the package test
suite).

## Feature-level texture learning in the generative pipeline

When training a generative texture model on real data, two mechanisms capture
feature-level structure:

- **Mediator panel (M)**: the feature-level Gaussian copula
  (`train_feature_texture()`) learns each feature's marginal distribution
  (empirical CDF or parametric family) and their cross-feature dependence
  (copula correlation matrix). During simulation, copula draws replace the
  parametric Gaussian noise in the mediator structural equations, so
  synthetic mediator data preserves the marginal shapes and correlation
  structure of the user's cohort.

- **Outcome (Y) and negative-control (W) panels**: feature-level residual
  correlation matrices are learned after residualizing on the sample-level
  exposure and covariates. During simulation, the noise terms in the outcome
  and negative-control equations are drawn from a multivariate normal with
  the learned correlation structure, so that conditional on the confounder
  $U$ the simulated features retain realistic cross-feature correlations.
  When feature correlation matrices are unavailable (e.g., panels not
  supplied or feature count mismatch), the noise defaults to independent
  draws.

The negative-control (W) panel is not learned from data via the copula; it
continues to use the pluggable `nc_model` with residual correlation matrices
injected as correlated noise.

The example data (`load_real_input_data(example = TRUE)`) includes
block-diagonal co-expression modules (within-block $r = 0.5$) so that the
learned correlations are non-trivial by default.

---

# Survival / time-to-event outcomes

ICONIC v0.9.4 extends the framework to support **Cox proportional hazards**
and **restricted mean survival time (RMST)** outcomes, motivated by case
studies where the outcome is overall survival (e.g., smoking $\to$ lung tumor
expression $\to$ lung cancer survival). All five total-effect estimators and
all eight mediation estimators have survival counterparts that share the same
2SPS (two-stage predictor substitution) structure: first-stage regressions
remain OLS because Z, M, and W are continuous; only the outcome stage switches
to `survival::coxph` (log-HR scale) or OLS on RMST pseudo-observations (time
scale).

## Two effect scales

| `effect_scale` | Outcome stage | NDE/NIE decomposition | Collapsible |
|---|---|---|---|
| `"loghr"` (default) | `coxph(Surv(time, event) ~ ...)` | Product $\alpha\beta$ on log-HR (approximate) | No |
| `"rmst"` | OLS on RMST pseudo-observations | Product $\alpha\beta$ on time scale (exact) | Yes |

The Cox log-HR is **non-collapsible**, so the product-of-coefficients
NIE $= \alpha_M \cdot \beta_M$ on the log-HR scale is an approximation. The
RMST scale (pseudo-observation OLS) provides the exact decomposition because
RMST differences are collapsible. RMST pseudo-observations are computed via
leave-one-out jackknife of Kaplan-Meier RMST (Graw et al. 2009), with the
truncation time $\tau$ defaulting to the 90th percentile of follow-up.

**COCA incompatibility:** COCA regresses W on Y (W ~ y + Z), placing the
outcome on the right-hand side — impossible with a `Surv` object. Both
`fit_coca_surv()` and `fit_coca_mediation_surv()` return `NA` with an
informative `"reason"` attribute for survival outcomes.

## Creating survival data

Pass `outcome_type = "survival"` with `surv_time` and `surv_event` instead of
`Y`:

```{r surv-data}
library(iconic)
set.seed(42)
dat <- generate_toy_data(n = 300, beta_Z = 0.25, alpha_M = 0.5,
                         beta_M = 0.3, conf_str = 0.8, phi = 0.8,
                         outcome_type = "survival", surv_event_frac = 0.6)
sdat <- iconic_data(Z = dat$Z, outcome_type = "survival",
                    surv_time = dat$surv_time, surv_event = dat$surv_event,
                    M = dat$M, G = dat$G[, 1], Gm = dat$Gm, W = dat$W)
sdat
```

## Estimating survival effects

```{r surv-estimate}
# Total effect on Cox log-HR scale
est_loghr <- iconic_estimate(sdat, effect_scale = "loghr")
est_loghr

# Total effect on RMST scale (collapsible, exact NDE/NIE decomposition)
est_rmst <- iconic_estimate(sdat, effect_scale = "rmst")
est_rmst
```

The log-HR estimates are interpretable as hazard ratios (exp(beta) = HR),
while RMST estimates are on a time scale (change in restricted mean survival
time per unit exposure). The NDE and NIE columns appear when mediation data
(M, Gm) is supplied.

## Survival simulation benchmark

The toy simulation supports `outcome_type = "survival"` in both
`generate_toy_data()` and `run_single_iteration()`. The linear predictor is
converted to time-to-event via an exponential proportional hazards model with
configurable baseline hazard (`surv_h0`), event fraction
(`surv_event_frac`), and censoring rate (`surv_censor_rate`). The true effects
(total, NDE, NIE) are on the Cox log-HR scale.

```{r surv-sim, eval=FALSE}
# Survival simulation: sweep confounding strength
# (analogous to sweep_param for continuous outcomes)
conf_grid <- c(0.2, 0.4, 0.6, 0.8, 1.0)
results <- data.frame()
for (cs in conf_grid) {
  for (i in 1:20) {
    dat <- generate_toy_data(n = 500, beta_Z = 0.25, alpha_M = 0.5,
                             beta_M = 0.3, conf_str = cs, phi = 0.8,
                             outcome_type = "survival", surv_event_frac = 0.6,
                             seed = 5000 + i)
    sdat <- iconic_data(Z = dat$Z, outcome_type = "survival",
                        surv_time = dat$surv_time, surv_event = dat$surv_event,
                        M = dat$M, G = dat$G[, 1], Gm = dat$Gm, W = dat$W)
    est <- iconic_estimate(sdat, effect_scale = "loghr")
    est$conf_str <- cs
    est$iter <- i
    results <- rbind(results, est)
  }
}
# Summarise bias vs confounding
library(dplyr)
summary <- results %>%
  group_by(conf_str, method) %>%
  summarise(NDE_bias = mean(NDE, na.rm = TRUE) - 0.25,
            NIE_bias = mean(NIE, na.rm = TRUE) - 0.15,
            .groups = "drop")
print(summary)
```

Sensitivity analysis (`iconic_sensitivity()`, `iconic_prospect()`) also
threads `outcome_type` and `effect_scale` through the simulation engine. The
GAN texture model is skipped for survival outcomes (it requires continuous Y);
the default texture is used instead.

---

# The full pipeline, end to end

```{r pipeline, eval=FALSE}
library(iconic)

# ── model selection workflow ──
# 1. Standardize data (optionally attach a pre-trained GAN)
data <- iconic_data(Z = my_exposure, Y = my_outcomes, M = my_mediator,
                    G = my_prs, Gm = my_eqtls, W1 = my_w1, W2 = my_w2,
                    covariates = my_covariates)

# 2. Diagnose
diag <- iconic_diagnose(data)

# 3. Estimate
est <- iconic_estimate(data, diagnosis = diag)

# 4. Infer confounding parameters (optional)
conf <- infer_confounding(data, diagnosis = diag, estimate = est)
print(conf)

# 5. Sensitivity (auto-trains hybrid GAN + copula from data;
#    or use confounding = "inferred")
sens <- iconic_sensitivity(data, diagnosis = diag,
                           confounding = "inferred", n_iter = 50)
print(sens)

# 6. Recommend
rec <- iconic_recommend(data, diagnosis = diag, estimate = est,
                        sensitivity = sens)
print(rec)

# ── Generative pipeline (standalone hybrid GAN + copula training) ──
input <- load_real_input_data(Z_matrix = my_gdm,      # features x samples
                              Y_matrix = my_expr,
                              covariates_df = my_meta)
gan   <- train_gan_on_real_data(input$gan_training_data, epochs = 300)

## Total-effect sensitivity
sens  <- gan_sensitivity(gan, conf_grid = c(0.2, 0.5, 0.8),
                         coverage_grid = c(0.3, 0.6, 1.0),
                         k_grid = c(1, 2, 3),
                         nc_model = "cpg", n_iter = 100, n_cores = 4)

recommend_estimator(sens)$overall
nc_validity_check(gan, k_grid = c(1, 2, 3), n_valid_controls = 2)$verdict

## Empirical NC validity screens (on your real fitted dataset)
# screen_Z <- nc_validity_screen(my_real_dat)
# screen_G <- nc_independence_check(my_real_dat)
# screen_Gm <- nc_independence_check_gm(my_real_dat)  # when phi > 0
# nc_completeness_check(my_real_dat, n_valid_controls = 2)

## Pleiotropy sensitivity
## (how robust to exclusion-restriction violations?)
pleio_sens <- gan_pleiotropy_sensitivity(gan,
    pleio_grid = c(0, 0.05, 0.10), conf_grid = c(0.2, 0.5, 0.8),
    n_iter = 50, n_features = 10)

## Mediation sensitivity (NDE/NIE under M-O confounding)
## Set phi > 0 to include the 2-stage MR estimator (IV2SLS2)
med_sens <- gan_mediation_sensitivity(gan,
    conf_grid = c(0.2, 0.5, 0.8), coverage_grid = c(0.3, 0.7, 1.0),
    mo_confounding = 0.8, phi = 0.8, n_iter = 100, n_cores = 4)
head(med_sens$summary[, c("method", "NDE_bias", "NIE_bias", "NIE_type1")])

plot_gan_diagnostics(gan, input$gan_training_data)
plot_sensitivity_heatmap(sens, metric = "rmse", method = "IV2SLS")
```

A ready-to-run version is in `inst/scripts/run_iconic_pipeline.R`.

---

# References

- Horn, J. L. (1965). A rationale and test for the number of factors in
  factor analysis. *Psychometrika*, 30(2), 179–185.
- Lipsitch, M., Tchetgen Tchetgen, E., & Cohen, T. (2010). Negative
  controls: a tool for detecting confounding and bias in observational
  studies. *Epidemiology*, 21(3), 383–388.
- Miao, W., Geng, Z., & Tchetgen Tchetgen, E. (2018). Identifying causal
  effects with proxy variables of an unmeasured confounder. *Biometrika*,
  105(4), 987–993.
- Tchetgen Tchetgen, E. (2014). The control outcome calibration approach
  for causal inference with unobserved confounding. *American Journal of
  Epidemiology*, 179(5), 633–640.
