---
title: "Genetic Instruments and Negative Controls from External Data"
author: "Sean T. Bresnahan and Charis Xiong"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Genetic Instruments and Negative Controls from External Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, comment = "#>"
)
set.seed(1)
library(iconic)
```

# Introduction

`iconic` identifies natural direct and indirect effects (NDE/NIE) in
observational omics data by combining *genetic instruments* (for the
exposure and, optionally, the mediator) with *negative controls* that proxy
the unmeasured confounders. In practice these inputs come from external
resources: GWAS summary statistics for exposure instruments, cis-eQTL
scans or transcriptome-wide association study (TWAS) weights for mediator
instruments, and high-dimensional omics panels (for example DNA
methylation) for negative-control construction.

This vignette demonstrates the helper functions that bridge those external
resources into an `iconic_data` object:

* `qc_gwas_sumstats()`, `build_prs_ldpred2()`, `score_pgs_panel()`, and
  `check_instrument_strength()` for exposure instruments;
* `call_cis_eqtls()` and `build_mediator_instruments()` for mediator
  instruments;
* `beta_to_m()`, `residualize_matrix()`, `build_w_pcs()`, and
  `apply_fusion_weights()` for negative-control panel construction;
* `as_iconic_data()` for importing a `SummarizedExperiment` directly.

All examples use small simulated inputs so the vignette builds quickly; on
real data the same calls apply unchanged.

# Installation

`iconic` is being submitted to Bioconductor. Once accepted, install the
release version with:

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("iconic")
```

The development version is available from GitHub:

```{r install-github, eval=FALSE}
BiocManager::install("sbresnahan/iconic")
```

# Exposure instruments from GWAS summary statistics

## Quality-control the summary statistics

`qc_gwas_sumstats()` standardises the column names of a GWAS summary
statistics table (recognising common aliases such as
`chromosome`/`base_pair_location`/`effect_allele`), and applies the usual
filters: missing values, extreme effect sizes, strand-ambiguous A/T and C/G
variants, and the standard-deviation-ratio check that compares the implied
phenotypic variance of each variant against the bulk distribution.

```{r qc-sumstats}
ss <- data.frame(
  chromosome = 1, base_pair_location = 1000 + 0:9 * 1000L,
  effect_allele = rep(c("A", "C"), 5), other_allele = rep(c("G", "T"), 5),
  beta = rnorm(10, 0, 0.05), standard_error = 0.01,
  effect_allele_frequency = runif(10, 0.1, 0.5)
)
out <- qc_gwas_sumstats(ss, n_eff = 50000)
out$qc
```

## Build a polygenic score with LDpred2

When the `bigsnpr` package is available, `build_prs_ldpred2()` matches the
summary statistics to an LD reference panel, runs LDpred2-auto (with a
moment-estimator fallback for the heritability), filters out
non-convergent chains, and can optionally score a target genotype panel.

```{r ldpred2, eval=requireNamespace("bigsnpr", quietly = TRUE)}
# Tiny fake genotype panel standing in for an LD reference
# (snp_fake genotypes are all-missing mocks; fill with random dosages)
fake <- bigsnpr::snp_fake(100, 500)
fake$genotypes[] <- rbinom(100 * 500, 2, 0.3)
fake$map$chromosome <- 1L
fake$map$physical.pos <- sort(sample(1:1e6, 500))
ss <- data.frame(
  chr = fake$map$chromosome, pos = fake$map$physical.pos,
  a0 = fake$map$allele2, a1 = fake$map$allele1,
  beta = rnorm(500, 0, 0.05), beta_se = 0.02, n_eff = 20000
)
prs <- build_prs_ldpred2(ss, ld_ref = fake, n_chains = 3,
                         burn_in = 50, num_iter = 50, verbose = FALSE)
str(prs$beta)
```

On real data the chain-convergence filter and the sd-ratio QC do the heavy
lifting; on this tiny fake panel the warning about non-convergent chains is
expected.

## Score a panel of pre-computed weights

If you already have effect weights (for example from the PGS Catalog or a
published score), `score_pgs_panel()` applies them to a dosage matrix
directly, matching variants by `chr:pos:ref:alt` key with automatic
strand-flip handling.

```{r score-panel}
dos <- matrix(rbinom(4 * 50, 2, 0.3), nrow = 4,
              dimnames = list(c("1:100:A:G", "1:200:C:T",
                                "1:300:G:A", "2:400:T:C"),
                              paste0("S", 1:50)))
wts <- data.frame(chr = c(1, 1, 2), pos = c(100, 300, 400),
                  a1 = c("A", "G", "T"), a0 = c("G", "A", "C"),
                  weight = c(0.1, -0.2, 0.05))
out <- score_pgs_panel(wts, dos)
head(out$score)
```

## Verify instrument strength

Whatever the source of the instrument, check the first-stage partial
F-statistic before relying on it. The conventional rule of thumb is
F >= 10.

```{r instrument-strength}
n <- 300
G <- rnorm(n)
X <- 0.3 * G + rnorm(n)
pcs <- matrix(rnorm(n * 3), n, 3, dimnames = list(NULL, paste0("PC", 1:3)))
check_instrument_strength(G, X, covariates = pcs)
```

# Mediator instruments

## Scan for cis-eQTLs

`call_cis_eqtls()` residualises expression and genotypes on covariates
(FWL) and runs a per-gene cis scan with Benjamini-Hochberg correction
within gene.

```{r cis-eqtls}
n <- 80
dos <- matrix(rbinom(200 * n, 2, 0.3), nrow = 200,
              dimnames = list(paste0("1:", 1:200, ":A:G"), NULL))
expr <- matrix(rnorm(20 * n), nrow = 20,
               dimnames = list(paste0("Gene", 1:20), NULL))
expr[1, ] <- expr[1, ] + 0.5 * dos[5, ]   # one true cis-eQTL
gp <- data.frame(gene = paste0("Gene", 1:20), chr = "1",
                 tss = seq(1, 191, by = 10))
sp <- data.frame(snp = rownames(dos), chr = "1", pos = 1:200)
hits <- call_cis_eqtls(expr, dos, gp, sp)
head(hits$best)
```

## Build a genetically predicted mediator (GReX-style)

`build_mediator_instruments()` fits an elastic-net cis-prediction model per
gene (via `glmnet`), keeps genes whose cross-validated out-of-fold
prediction passes quality gates, and returns the predicted mediator matrix
`Gm` in the genes x samples orientation expected by `iconic_data()`.

```{r mediator-instruments, eval=requireNamespace("glmnet", quietly = TRUE)}
set.seed(42)
n <- 150; p <- 80
dos <- matrix(rbinom(p * n, 2, 0.3), nrow = p,
              dimnames = list(paste0("1:", 1:p, ":A:G"),
                              paste0("S", 1:n)))
expr <- matrix(rnorm(4 * n), nrow = 4,
               dimnames = list(paste0("Gene", 1:4), paste0("S", 1:n)))
expr[1, ] <- expr[1, ] + 0.9 * scale(dos[5, ]) + 0.9 * scale(dos[15, ])
gp <- data.frame(gene = paste0("Gene", 1:4), chr = "1",
                 tss = c(5, 25, 45, 65))
sp <- data.frame(snp = rownames(dos), chr = "1", pos = 1:p)
fit <- build_mediator_instruments(expr, dos, gp, sp, seed = 1)
fit$qc
```

Only Gene1, which carries the two planted cis-eQTLs, passes the
cross-validation gates.

# Negative controls from high-dimensional panels

## Methylation beta to M-values

Methylation beta values are heteroscedastic at the extremes;
`beta_to_m()` applies the logit transform with clipping.

```{r beta-to-m}
b <- matrix(runif(200, 0.01, 0.99), nrow = 20,
            dimnames = list(paste0("cg", 1:20), paste0("S", 1:10)))
m <- beta_to_m(b)
range(m)
```

## Residualise on covariates

Before extracting negative-control factors, remove technical covariates
(for example batch) from the panel.

```{r residualize}
x <- matrix(rnorm(100 * 40), nrow = 100,
            dimnames = list(paste0("f", 1:100), paste0("S", 1:40)))
batch <- factor(rep(c("A", "B"), each = 20))
cv <- model.matrix(~ batch)[, -1, drop = FALSE]
xr <- residualize_matrix(x, cv)
cor(as.numeric(xr[1, ]), as.numeric(cv))
```

## Extract negative-control factors with PCA

`build_w_pcs()` computes the top principal components of the (residualised)
panel — the `W` matrix used by the proximal and COCA estimators.

```{r build-w}
x <- matrix(rnorm(500 * 60), nrow = 500,
            dimnames = list(paste0("f", 1:500), paste0("S", 1:60)))
w <- build_w_pcs(x, n_pcs = 5)
dim(w$W)
w$variance_explained
```

## Apply FUSION-style TWAS weights

Pre-computed transcriptomic weights (FUSION `.wgt.RDat` files, or an
in-memory list with the same structure) can be applied to a dosage panel to
build a genetically predicted expression panel, which can itself serve as a
negative-control or mediator-instrument panel.

```{r fusion-weights}
dos <- matrix(rbinom(10 * 50, 2, 0.3), nrow = 10,
              dimnames = list(paste0("rs", 1:10), paste0("S", 1:50)))
mk_wgt <- function(snps, w) {
  list(wgt.matrix = matrix(w, ncol = 1, dimnames = list(NULL, "enet")),
       snps = data.frame(V2 = snps),
       cv.performance = matrix(0.2, nrow = 1,
                               dimnames = list("rsq", "enet")))
}
wlist <- list(GeneA = mk_wgt(paste0("rs", 1:5), rep(0.1, 5)),
              GeneB = mk_wgt(paste0("rs", 6:10), rep(-0.2, 5)))
pos <- data.frame(ID = c("GeneA", "GeneB"), WGT = c("a.wgt.RDat", "b.wgt.RDat"))
out <- apply_fusion_weights(dos, pos = pos, weights = wlist)
dim(out$predicted)
```

# Importing a SummarizedExperiment

For Bioconductor-native workflows, `as_iconic_data()` dispatches on
`SummarizedExperiment`: the primary assay becomes the outcome panel `Y`, an
optional second assay becomes the mediator panel `M`, and `colData`
columns supply the exposure, instruments, negative controls, covariates,
and survival endpoints.

```{r summarized-experiment, eval=requireNamespace("SummarizedExperiment", quietly = TRUE)}
se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(expr = matrix(rnorm(20 * 60), 20, 60,
                              dimnames = list(paste0("gene", 1:20),
                                              paste0("S", 1:60)))),
  colData = S4Vectors::DataFrame(
    bmi = rnorm(60), prs = rnorm(60),
    nc1 = rnorm(60), nc2 = rnorm(60), age = rnorm(60))
)
data <- as_iconic_data(se, assay = "expr", exposure = "bmi",
                       instrument = "prs",
                       negative_controls = c("nc1", "nc2"),
                       covariates = "age")
print(data)
```

From here the standard workflow applies: `iconic_diagnose()`,
`iconic_estimate()`, and the sensitivity machinery described in
`vignette("iconic-walkthrough")`.

# Session information

```{r sessionInfo}
sessionInfo()
```
