---
title: "Diagnosing Estimator Eligibility"
author: "Sean T. Bresnahan and Charis Xiong"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Diagnosing Estimator Eligibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, comment = "#>"
)
set.seed(1)
library(iconic)
```

# What diagnosis checks

Before fitting any estimator, `iconic_diagnose()` examines your data and
reports which of the eight ICONIC estimators are *eligible* — that is, which
satisfy the assumptions they require to be valid. These assumptions are
not directly testable (the confounder $U$ is unobserved); diagnosis checks
their empirically observable projections. Diagnosis checks four things:

1. **Instrument strength.** The partial $F$-statistic for each genetic
   instrument ($G_1$ for the exposure, $G_m$ for the mediator) against the
   conventional weak-instrument threshold ($F \geq 10$).
2. **Negative-control validity.** Three screens — A1 (NCs associate with
   the exposure), A2 (NCs are independent of the instrument given
   covariates), and A2' (NCs associate with the outcome given the exposure)
   — assessed via partial correlations with BH-FDR control.
3. **Completeness.** Whether the number of valid negative controls is
   sufficient for the proximal estimators (PGC, PGC2, PGC2Gm), which
   require $\dim(W_\text{valid}) \geq k$ (the number of latent confounders).
4. **Eligibility table.** A per-estimator verdict: eligible, ineligible
   (with reason), or not applicable.

# A minimal example

We simulate a small mediation panel with a genetic instrument and negative
controls, then run the diagnosis.

```{r diagnose-data}
dat <- generate_toy_data(
  n = 100, n_features = 5, seed = 1,
  phi = 0.8, omega_1 = 0.7, omega_2 = 0.7
)

idat <- iconic_data(
  X = dat$X, Y = dat$Y, M = dat$M,
  G = dat$G, Gm = dat$Gm,
  W1 = dat$W1, W2 = dat$W2,
  covariates = dat$synthetic_data
)
```

```{r diagnose-run}
diag <- iconic_diagnose(idat)
diag$eligibility
```

Each row reports whether the estimator is eligible and, if not, why. For
example, `IV2SLS` requires a valid exposure instrument ($F \geq 10$), while
`PGC` additionally requires negative-control completeness.

# Instrument strength

```{r diagnose-instrument}
diag$instrument_strength
```

The report gives the partial $F$-statistic for $G_1$ (exposure instrument)
and $G_m$ (mediator instrument). Estimators that rely on an instrument
(`IV2SLS`, `IV2SLS2`, `PGC2Gm`) are eligible only when the relevant
$F$-statistic meets the threshold.

# Negative-control validity

```{r diagnose-nc}
diag$nc_validity
```

The NC validity screen tests the empirically observable projections of
three identifying assumptions for each negative-control feature:

- **A1**: $W \perp X \mid C, U$ (the NC is not affected by the exposure
  once the confounder is accounted for). The screen tests the observable
  projection: association of $W$ with $X$ given $C$.
- **A2**: $W \perp G_1 \mid C, U$ (the NC is not an instrument itself).
  The screen tests the observable projection: association of $W$ with
  $G_1$ given $C$.
- **A2'**: $W \perp G_m \mid C, U$ (the NC is not a mediator instrument).
  The screen tests the observable projection: association of $W$ with
  $G_m$ given $C$.

Features that fail a screen are flagged, and the completeness count is
reduced accordingly. The `COCA` estimator is exempt from the A2 screen
because it calibrates through the $W \sim Y + X$ ratio and does not require
$W \perp G_1$.

# Proceeding without instruments or controls

If you supply neither genetic instruments nor negative controls, only the
`UNADJ` estimator (the confounded OLS reference) is eligible. By default,
`iconic_diagnose()` proceeds with a message:

```{r diagnose-no-proxy, eval=FALSE}
dat0 <- generate_toy_data(n = 100, n_features = 5, seed = 1)
idat0 <- iconic_data(X = dat0$X, Y = dat0$Y, covariates = dat0$synthetic_data)
diag0 <- iconic_diagnose(idat0)
```

Set `allow_no_proxy = FALSE` to make this an error instead, which is useful
in pipelines where you want to enforce that identification assumptions are
met before proceeding.

# Further reading

- **Walkthrough vignette**: `vignette("iconic-walkthrough")` — the full
  model-selection workflow.
- **Function reference**: `?iconic_diagnose` for all arguments, including
  `min_f`, `fdr_level`, and the per-estimator NC screens.
