---
title: "Sensitivity Analysis and Confounding Inference"
author: "Sean T. Bresnahan and Charis Xiong"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Sensitivity Analysis and Confounding Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, comment = "#>"
)
set.seed(1)
library(iconic)
has_torch <- check_torch_setup()
```

# What sensitivity analysis does

Even after diagnosis declares an estimator eligible, its validity rests on
the assumption that the genetic instrument is exogenous — that the
instrument-confounder correlations $\rho_{G1}$ and $\rho_{G2}$ are zero.
These correlations cannot be estimated from data (the confounders are
unobserved), so `iconic_sensitivity()` sweeps them over a plausible grid
and reports how each estimator's bias degrades.

The output reports, for each estimator:

- **Maximum absolute bias** across the sensitivity surface.
- **Tipping points**: the first $\rho_{G2}$ at $\rho_{G1} = 0$ where
  $|\text{bias}| > 0.10$.
- **Texture source**: whether the synthetic data used the generative
  texture model (GAN + copula) or the parametric benchmark generator.

# A minimal example

We simulate a small mediation panel, train a texture model briefly, and
run the sensitivity sweep.

```{r sens-data}
dat <- generate_toy_data(
  n = 200, n_features = 10, seed = 1,
  phi = 0.8, omega_1 = 0.7, omega_2 = 0.7
)

idat <- iconic_data(
  X = dat$X, Y = dat$Y, M = dat$M,
  G = dat$G, Gm = dat$Gm,
  W1 = dat$W1, W2 = dat$W2,
  covariates = dat$synthetic_data
)
```

```{r sens-gan, eval=has_torch}
# Train a texture model briefly and attach it to the iconic_data object.
input <- load_real_input_data(
  X_matrix = matrix(dat$X, nrow = 1),
  Y_matrix = t(dat$Y), M_matrix = t(dat$M),
  W_matrix = t(dat$W), covariates_df = dat$synthetic_data
)

idat$trained_gan <- train_gan_on_real_data(
  input$gan_training_data,
  feature_correlations = input$feature_correlations,
  feature_texture = input$feature_texture,
  epochs = 5, seed = 1, verbose = FALSE
)
```

```{r sens-no-torch, eval=!has_torch, echo=FALSE, results="asis"}
cat(
"The sensitivity-sweep chunks in this vignette are skipped because torch ",
"is not available. Install torch with `install.packages(\"torch\")` and ",
"run `torch::install_torch()` to run them."
)
```

```{r sens-run, eval=has_torch}
sens <- iconic_sensitivity(idat, n_iter = 3)
sens
```

# Inferring confounding from the data

`infer_confounding()` estimates the held-fixed confounding parameters from
your data, supplying a `confounding = "inferred"` mode for
`iconic_sensitivity()` and `iconic_prospect()`.

```{r sens-infer}
diag <- iconic_diagnose(idat)
est <- iconic_estimate(idat, diagnosis = diag)
conf <- infer_confounding(idat, diag, est)
conf
```

The inferred parameters include:

- **Confounding strength** ($\delta$): from the gap between unadjusted OLS
  and IV2SLS estimates.
- **Mediator-outcome confounding** ($\delta_{mo}$): from the gap between
  IV2SLS and IV2SLS2 natural indirect effects.
- **Negative-control coverage** ($\omega_1$, $\omega_2$): from the
  $R^2$ of each NC feature regressed on the outcome residual.
- **Number of latent confounders** ($k$): via parallel analysis on the
  residualized outcome correlation matrix.

Parameters that cannot be inferred (the instrument-confounder correlations
$\rho_{G1}$, $\rho_{G2}$) fall back to defaults with a warning and are
always swept.

# Inferred vs. default confounding

```{r sens-inferred, eval=has_torch}
sens_inf <- iconic_sensitivity(idat, n_iter = 3, confounding = "inferred")
sens_inf
```

The `confounding = "default"` mode uses fixed defaults
($\delta_{mo} = 0.8$, $\omega_1 = \omega_2 = 0.7$, $\phi = 0.8$) and is the
recommended starting point. The `confounding = "inferred"` mode calibrates
to your data but uses estimator validity (e.g., that IV2SLS is unbiased) to
set up a benchmark whose purpose is to test estimator validity — so
inferred values should be read as best-case calibrations under the stated
assumptions.

# Further reading

- **Texture model vignette**: `vignette("iconic-texture-model")` — how the
  GAN + copula texture model is trained and why it matters for sensitivity.
- **Function reference**: `?iconic_sensitivity` for the full argument list,
  including the $\rho_{G1} \times \rho_{G2}$ grid and the `n_iter` control.
